\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-20 writing/manuscript.Rmd
added to modeling $R$ and spatial needs

In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

@@

In probability and statistics, random variables are defined by specifying the distribution the variable follows. Since the variable is random, it can take multiple values (minimum 2), each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.

@@

The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.

@@

A random variable's variance is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

@@

In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation

@@

@nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

@@

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident. ***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***

@@

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs]. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

@@

Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs. These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@examples].

Recent increases in the frequency and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018]. When possible, increasing the frequency of extreme events to reduce their frequency (such as with controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].

@@

In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable. Thus, we expect $\var(R)$ to have a stronger effect on resource abundance when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.

@@

\noindent

This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref]. Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].

@@

The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].

Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples]. Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].

@@

The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism (i.e. approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs]. Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples]. Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].

@@

The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses. We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for. In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings. Additionally, the analyses can be expanded to test for commonalities and differences in the behaviors between members of the same population or species, as well as between multiple distinct populations or species.

@@

Location-scale models are a class of statistical models that allow one to model the mean and variance of the response rather than only modeling the mean.

* location-scale models
* `mgcv`, `brms`
* choosing the best family
* choosing the right complexity of each term (maybe that's why @nilsen_can_2005 found contrasting results)

@@

* importance of AKDE
* Introduce moving window idea

Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017]. However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...

AKDE is good for range-resident animals

In the case of animals whose spatial needs change over their lifetimes [arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017], one may define a HR as a function of time with distinct HRs for each period or as a continuous function of time. In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available. However, when an animal's spatial needs change over time, models which assume range residency are not appropriate. Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results. Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].

@@

Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values. Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.

@@

Animal's spatial needs are often estimated using all available tracking data. Doing so maximizes the (effective) sample size, but it also assumes that during the tracking period the animal's movement behavior did not change and that its home range did not shift either. Although these assumptions may be valid for relatively short tracking periods (and often necessary for reasonable estimates of spatial needs), few animals will behave consistently during the tracking period and use the same space every day. ***Most animals (if not all) use their space selectively.*** *add examples with humans moving to different regions?*

@@

@@

>>>>

\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

@@

\noindent In probability and statistics, random variables are defined by specifying the distribution the variable follows. Since the variable is random, it can take multiple values (minimum 2), each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.

@@

\noindent The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.

@@

\noindent A random variable's variance is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

@@

\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation

@@

\noindent @nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

\noindent To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

@@

\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident. ***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***

@@

\noindent Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017; @yao_emergence_2022], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs]. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

@@

Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs. These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].

Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018]. Managing the intensity of and frequency of extreme events (e.g., controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].

@@

\noindent In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable. Thus, we expect $\var(R)$ to have a stronger effect on resource abundance when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.

@@

\noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref]. Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].

@@

\noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019]. Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples]. Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].

@@

@@

\noindent The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses. We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for. In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings. Additionally, the analyses can be expanded to test for commonalities and differences in the behaviors between members of the same population or species, as well as between multiple distinct populations or species.

@@

\noindent Location-scale models [@stasinopoulos_generalized_2007, @rigby_generalized_2005] are a class of statistical models that us to estimate changes in $\e(R)$ and $\var(R)$ while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly-used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017] for location and scale (GAMLSs). Currently, the `mgcv` package allows one to fit location-scale models with a variety distribution families, including Gaussian, scaled t, Gamma, Tweedie, Gumbel families.

The Gaussian family of distributions is the most flexible, since the mean and standard deviation parameters are assumed to be independent, and the response can be either positive or negative. The scaled t family is similar to the Gaussian family, but it is best for high-variance data with heavy tails. However, the two distributions' flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data, count data, proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).

The Gamma location-scale family is best for strictly positive responses, including elemental compositions (e.g., carbon to nitrogen ratio), total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day). The Tweedie family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey abundance. Lastly, the Gumbel distribution is suitable for modeling the extremes of a distribution, such as maximum or minimum daily temperatures.

If one is interested in using other families of distributions, including zero-inflated and hurdle models, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of the family's parameters using a fully Bayesian approach [@burkner_advanced_2018]. In this paper, we use the `mgcv` package because although `brms` provides more flexibility and a greater choice of families, its computational costs can be prohibitive for large datasets, particularly for location-scale models.

Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully. Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and not monotonic, we suggest using a GAM rather than a GLM. However, the complexity of the spatio-temporal terms should be chosen carefully, particularly the terms for the mean (i.e., location). Excessively wiggly mean terms will cause the variance to be under-estimated and under-fit, and an excessively smooth mean terms will inflate the variance term, so an improperly defined model can lead to inaccurate estimates. Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt should be sufficient. Additionally, REstricted Marginal Likelihood [a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results. We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals. Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies.

@@

\noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017]. Since all models that can be currently fit via the `ctmm` package [@fleming_ctmm_2021] assume the animal to be range-resident and move with a single, consistent behavior, we suggest estimating an animal's spatial needs using a **sliding window** approach.

**here**

Let $w$ indicate a fixed window width.

In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available. However, when an animal's spatial needs change over time, models which assume range residency are not appropriate. Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results. Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].

@@

\noindent Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values. Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.

@@

\noindent Animal's spatial needs are often estimated using all available tracking data. Doing so maximizes the (effective) sample size, but it also assumes that during the tracking period the animal's movement behavior did not change and that its home range did not shift either. Although these assumptions may be valid for relatively short tracking periods (and often necessary for reasonable estimates of spatial needs), few animals will behave consistently during the tracking period and use the same space every day. ***Most animals (if not all) use their space selectively.*** *add examples with humans moving to different regions?*

@@

Additional info is in appendix 2

we use NDVI [@ref] as a proxy for resource abundance.

@@

\noindent text
